---
title: "Coursework 2"
subtitle: "MA22019 Introduction to Data Science"
format: gfm
editor: visual
---

<!--- DO NOT DELETE THIS LINE --->

<!--- METADATA ANCHOR: MA22019-CW2-2026 --->

# Data Preparation Notes

<!--- DO NOT DELETE THIS LINE - DATA PREPARATION NOTES ANCHOR --->

Briefly document the main data-handling decisions that matter for your analysis. This section may be short prose or one small table.

```{r}
#| label: data-preparation-code
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(knitr)

wards <- read_sf("data/city_wards.shp")
profiles <- read_csv("data/ward_profiles.csv" , show_col_types = FALSE)
incidents <- read_csv("data/incident_reports.csv" , show_col_types = FALSE)

#NAs in datasets
na_summary <- tibble(dataset = c("incidents", "ward profiles", "ward polygons"),
  missing_values = c(sum(is.na(incidents)),sum(is.na(profiles)),sum(is.na(wards))))

#conversion and check in spatial
incidents_sf <- incidents %>% 
  st_as_sf( coords = c("x","y"), crs = st_crs(wards) )

#creating recorded and spatial wards left join
incidents_joined <- incidents_sf %>%
  st_join( wards %>% select(ward_id , ward_name) , left = TRUE ) %>% 
  rename( ward_name_recorded = ward_name.x , ward_name_spatial = ward_name.y)

#number of rows where recorded and spatial dont match
discrepancies <- incidents_joined %>% 
  st_drop_geometry() %>% 
  filter(ward_name_recorded != ward_name_spatial) %>% 
  nrow()

#number of incidents in each spatial ward
incident_counts <-  incidents_joined %>% 
  st_drop_geometry() %>%
  count(ward_id, ward_name_spatial, name = "n_incidents")
  
#profiles cleaning 
profiles_clean <- profiles %>%
  mutate(rental_share = parse_number(rental_share),
         ward_name = str_replace_all(ward_name, "-" , " "))

#final data set contain - count and profiles all by spatial crs

wards_data <- wards %>% 
  left_join(
    incident_counts %>% select(ward_id, n_incidents),
    by = "ward_id"
  ) %>%
  mutate(n_incidents = replace_na(n_incidents, 0)) %>%
  left_join(
    profiles_clean %>% select(-ward_name),
    by = "ward_id"
  ) %>% 
  mutate( area_km2 = as.numeric(st_area(geometry))/1000000,
          incident_density =  n_incidents / area_km2)


na_summary
discrepancies
```

The incident records are point pattern data, with each row giving an event location. I converted the incident coordinates to an sf point object using the ward boundary CRS and spatially joined the points to ward polygons. This allowed the recorded ward names to be checked against the coordinate-based ward assignment. Ward-level incident counts were then calculated using the spatially assigned wards. I converted rental_share from percentage text to numeric and joined ward-profile variables by ward_id, avoiding reliance on ward names because some profile names used hyphens. The ward profile file had 6 missing values: 3 in transport_access and 3 in listed_building_share which were retained. I also calculated ward area and incident density to compare wards of different sizes.

# Task 1

<!--- DO NOT DELETE THIS LINE - TASK 1 ANCHOR --->

Identify and investigate the most important spatial patterns in the incident data.

```{r}
#| label: task-1-code
# Write your Task 1 code here.
```

Write your Task 1 answer here.

# Task 2

<!--- DO NOT DELETE THIS LINE - TASK 2 ANCHOR --->

Use the available neighbourhood data to develop and justify possible explanations for the patterns you identified.

```{r}
#| label: task-2-code
# Write your Task 2 code here.
```

Write your Task 2 answer here.

# Task 3

<!--- DO NOT DELETE THIS LINE - TASK 3 ANCHOR --->

Make a small number of evidence-based recommendations for priority areas or actions.

```{r}
#| label: task-3-code
# Write your Task 3 code here.
```

Write your Task 3 answer here.

# References

<!--- DO NOT DELETE THIS LINE - REFERENCES ANCHOR --->

Add references only if needed.

```{r}
#| label: word-count-check
#| echo: false
#| message: false
#| warning: false

resolve_word_count_input <- function() {
  current_input <- tryCatch(
    knitr::current_input(),
    error = function(e) NA_character_
  )

  if (!is.na(current_input) && nzchar(current_input) && file.exists(current_input)) {
    return(current_input)
  }

  if (file.exists("coursework_02.qmd")) {
    return("coursework_02.qmd")
  }

  stop(
    paste(
      "Could not locate coursework_02.qmd.",
      "Render the document from inside the coursework project folder."
    )
  )
}

extract_chunk_captions <- function(chunk_lines) {
  caption_lines <- grep(
    "^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:",
    chunk_lines,
    value = TRUE
  )

  if (length(caption_lines) == 0) {
    return(character(0))
  }

  trimws(sub("^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:\\s*", "", caption_lines))
}

count_policy_words <- function(text) {
  if (length(text) == 0) {
    return(0L)
  }

  text <- paste(text, collapse = " ")
  text <- gsub("`[^`]*`", " ", text, perl = TRUE)
  text <- gsub("\\[[^\\]]*\\]\\([^\\)]*\\)", " ", text, perl = TRUE)
  text <- gsub("https?://\\S+", " ", text, perl = TRUE)
  text <- gsub("[*_~#>]", " ", text, perl = TRUE)

  tokens <- regmatches(
    text,
    gregexpr("[[:alnum:]]+(?:[-'][[:alnum:]]+)*", text, perl = TRUE)
  )[[1]]

  if (length(tokens) == 1 && identical(tokens, character(0))) {
    return(0L)
  }

  length(tokens)
}

count_prose_words <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)

  yaml_bounds <- grep("^---\\s*$", lines)
  if (length(yaml_bounds) >= 2) {
    lines <- lines[-seq.int(yaml_bounds[1], yaml_bounds[2])]
  }

  reference_heading <- grep("^#\\s+References\\s*$", lines)
  if (length(reference_heading) > 0) {
    lines <- lines[seq_len(reference_heading[1] - 1)]
  }

  prose_lines <- character(0)
  caption_lines <- character(0)
  in_chunk <- FALSE
  current_chunk <- character(0)

  for (line in lines) {
    if (grepl("^```", line)) {
      if (!in_chunk) {
        in_chunk <- TRUE
        current_chunk <- character(0)
      } else {
        caption_lines <- c(caption_lines, extract_chunk_captions(current_chunk))
        in_chunk <- FALSE
        current_chunk <- character(0)
      }
      next
    }

    if (in_chunk) {
      current_chunk <- c(current_chunk, line)
      next
    }

    trimmed <- trimws(line)
    if (
      trimmed == "" ||
      grepl("^<!--", trimmed) ||
      grepl("^#+\\s+", trimmed)
    ) {
      next
    }

    prose_lines <- c(prose_lines, line)
  }

  total_words <- count_policy_words(prose_lines) + count_policy_words(caption_lines)
  total_words
}

true_word_count <- count_prose_words(resolve_word_count_input())
word_limit <- 1000L
status_text <- if (true_word_count <= word_limit) "under" else "over"
delta <- abs(word_limit - true_word_count)

cat(
  sprintf(
    "**Prose Word Count:** %d words (%d words %s the %d-word limit)",
    true_word_count,
    delta,
    status_text,
    word_limit
  )
)
```
