---
title: "Coursework 2"
subtitle: "MA22019 Introduction to Data Science"
format: gfm
editor: visual
---

<!--- DO NOT DELETE THIS LINE --->
<!--- METADATA ANCHOR: MA22019-CW2-2026 --->

# Data Preparation Notes

<!--- DO NOT DELETE THIS LINE - DATA PREPARATION NOTES ANCHOR --->

Briefly document the main data-handling decisions that matter for your
analysis. This section may be short prose or one small table.

```{r}
#| label: data-preparation-code
# Write your data-preparation code here.
```

Write your `Data Preparation Notes` answer here.

# Task 1

<!--- DO NOT DELETE THIS LINE - TASK 1 ANCHOR --->

Identify and investigate the most important spatial patterns in the incident
data.

```{r}
#| label: task-1-code
# Write your Task 1 code here.
```

Write your Task 1 answer here.

# Task 2

<!--- DO NOT DELETE THIS LINE - TASK 2 ANCHOR --->

Use the available neighbourhood data to develop and justify possible
explanations for the patterns you identified.

```{r}
#| label: task-2-code
# Write your Task 2 code here.
```

Write your Task 2 answer here.

# Task 3

<!--- DO NOT DELETE THIS LINE - TASK 3 ANCHOR --->

Make a small number of evidence-based recommendations for priority areas or
actions.

```{r}
#| label: task-3-code
# Write your Task 3 code here.
```

Write your Task 3 answer here.

# References

<!--- DO NOT DELETE THIS LINE - REFERENCES ANCHOR --->

Add references only if needed.

```{r}
#| label: word-count-check
#| echo: false
#| message: false
#| warning: false

resolve_word_count_input <- function() {
  current_input <- tryCatch(
    knitr::current_input(),
    error = function(e) NA_character_
  )

  if (!is.na(current_input) && nzchar(current_input) && file.exists(current_input)) {
    return(current_input)
  }

  if (file.exists("coursework_02.qmd")) {
    return("coursework_02.qmd")
  }

  stop(
    paste(
      "Could not locate coursework_02.qmd.",
      "Render the document from inside the coursework project folder."
    )
  )
}

extract_chunk_captions <- function(chunk_lines) {
  caption_lines <- grep(
    "^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:",
    chunk_lines,
    value = TRUE
  )

  if (length(caption_lines) == 0) {
    return(character(0))
  }

  trimws(sub("^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:\\s*", "", caption_lines))
}

count_policy_words <- function(text) {
  if (length(text) == 0) {
    return(0L)
  }

  text <- paste(text, collapse = " ")
  text <- gsub("`[^`]*`", " ", text, perl = TRUE)
  text <- gsub("\\[[^\\]]*\\]\\([^\\)]*\\)", " ", text, perl = TRUE)
  text <- gsub("https?://\\S+", " ", text, perl = TRUE)
  text <- gsub("[*_~#>]", " ", text, perl = TRUE)

  tokens <- regmatches(
    text,
    gregexpr("[[:alnum:]]+(?:[-'][[:alnum:]]+)*", text, perl = TRUE)
  )[[1]]

  if (length(tokens) == 1 && identical(tokens, character(0))) {
    return(0L)
  }

  length(tokens)
}

count_prose_words <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)

  yaml_bounds <- grep("^---\\s*$", lines)
  if (length(yaml_bounds) >= 2) {
    lines <- lines[-seq.int(yaml_bounds[1], yaml_bounds[2])]
  }

  reference_heading <- grep("^#\\s+References\\s*$", lines)
  if (length(reference_heading) > 0) {
    lines <- lines[seq_len(reference_heading[1] - 1)]
  }

  prose_lines <- character(0)
  caption_lines <- character(0)
  in_chunk <- FALSE
  current_chunk <- character(0)

  for (line in lines) {
    if (grepl("^```", line)) {
      if (!in_chunk) {
        in_chunk <- TRUE
        current_chunk <- character(0)
      } else {
        caption_lines <- c(caption_lines, extract_chunk_captions(current_chunk))
        in_chunk <- FALSE
        current_chunk <- character(0)
      }
      next
    }

    if (in_chunk) {
      current_chunk <- c(current_chunk, line)
      next
    }

    trimmed <- trimws(line)
    if (
      trimmed == "" ||
      grepl("^<!--", trimmed) ||
      grepl("^#+\\s+", trimmed)
    ) {
      next
    }

    prose_lines <- c(prose_lines, line)
  }

  total_words <- count_policy_words(prose_lines) + count_policy_words(caption_lines)
  total_words
}

true_word_count <- count_prose_words(resolve_word_count_input())
word_limit <- 1000L
status_text <- if (true_word_count <= word_limit) "under" else "over"
delta <- abs(word_limit - true_word_count)

cat(
  sprintf(
    "**Prose Word Count:** %d words (%d words %s the %d-word limit)",
    true_word_count,
    delta,
    status_text,
    word_limit
  )
)
```
