---
title: "Coursework 2"
subtitle: "MA22019 Introduction to Data Science"
format: gfm
editor: visual
---

<!--- DO NOT DELETE THIS LINE --->

<!--- METADATA ANCHOR: MA22019-CW2-2026 --->

# Data Preparation Notes

<!--- DO NOT DELETE THIS LINE - DATA PREPARATION NOTES ANCHOR --->

Briefly document the main data-handling decisions that matter for your analysis. This section may be short prose or one small table.

```{r}
#| label: data-preparation-code
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(knitr)

wards <- read_sf("data/city_wards.shp")
profiles <- read_csv("data/ward_profiles.csv" , show_col_types = FALSE)
incidents <- read_csv("data/incident_reports.csv" , show_col_types = FALSE)

#NAs in datasets
na_summary <- tibble(dataset = c("incidents", "ward profiles", "ward polygons"),
  missing_values = c(sum(is.na(incidents)),sum(is.na(profiles)),sum(is.na(wards))))
  
#profiles cleaning 
profiles_clean <- profiles %>%
  mutate(rental_share = parse_number(rental_share),
         ward_name = str_replace_all(ward_name, "-" , " "))

na_summary

```
I loaded the ward boundary polygons, ward profile data and incident reports. The incident and ward polygon files had no missing values, while the ward profile file had 6 missing values; these were retained and handled using available cases where relevant. I converted rental_share from percentage text to a numeric variable. I also standardized ward names in the profile file by replacing hyphens with spaces, although later joins use ward_id rather than ward_name to avoid relying on name formatting.

# Task 1

<!--- DO NOT DELETE THIS LINE - TASK 1 ANCHOR --->

Identify and investigate the most important spatial patterns in the incident data.

```{r}
#| label: task-1-code
#| message: false
#| warning: false

library(spatstat)

#conversion and check in spatial
incidents_sf <- incidents %>% 
  st_as_sf( coords = c("x","y"), crs = st_crs(wards) )

#creating recorded and spatial wards left join
incidents_joined <- incidents_sf %>%
  st_join( wards %>% select(ward_id , ward_name) , left = TRUE ) %>% 
  rename( ward_name_recorded = ward_name.x , ward_name_spatial = ward_name.y)

#number of rows where recorded and spatial dont match
discrepancies <- incidents_joined %>% 
  st_drop_geometry() %>% 
  filter(ward_name_recorded != ward_name_spatial) %>% 
  nrow()

#number of incidents in each spatial ward
incident_counts <-  incidents_joined %>% 
  st_drop_geometry() %>%
  count(ward_id, ward_name_spatial, name = "n_incidents")

wards_incident <- wards %>% 
  left_join(incident_counts %>% select(ward_id, n_incidents),by = "ward_id") %>%
  mutate(n_incidents = replace_na(n_incidents, 0),
         area_km2 = as.numeric(st_area(geometry))/1000000,
         incident_density =  n_incidents / area_km2)

#kernel-smoothed intensity for ppp and quadrat with 4cm
wards_window <- as.owin(st_union(wards))
incident_coords = st_coordinates(incidents_sf)

incident_ppp = ppp( x = incident_coords[,1] , y =incident_coords[,2] , window = wards_window)

incident_quadrat6 <- quadratcount( incident_ppp, nx=6 , ny=6 )
plot(incident_ppp,pch = 16,cex = 0.25,main = "Quadrat counts of reported incidents")
plot(incident_quadrat6, add = TRUE)
plot(st_geometry(wards),add = TRUE,lwd = 0.5)

incident_intensity = density.ppp( x = incident_ppp , edge= TRUE)
plot(incident_intensity,main = "Kernel-smoothed intensity of reported incidents")
plot(st_geometry(wards),add = TRUE,lwd = 0.5)
plot(incident_ppp, add = TRUE, pch = 16,cex = 0.25)

top_wards <- wards_incident %>% 
  st_drop_geometry() %>% 
  arrange(desc(n_incidents)) %>% 
  select(ward_id,ward_name, n_incidents, incident_density) %>% 
  slice_head(n=10)
top_wards

total_incidents = sum(wards_incident$n_incidents)

top_10_share <- wards_incident %>%
  st_drop_geometry() %>% 
  arrange(desc(n_incidents)) %>% 
  slice_head(n =10) %>% 
  summarise( share = (sum(n_incidents)/total_incidents) *100)
top_10_share

top_5_share <-  wards_incident %>% 
  st_drop_geometry() %>% 
  arrange(desc(n_incidents)) %>% 
  slice_head(n=5) %>% 
  summarise ( share = (sum(n_incidents)/total_incidents)*100)
top_5_share 

```
The incident records were treated as point pattern data. A 6 × 6 quadrat count showed strong variation across the city, with the largest counts in the north-east/eastern cells and several southern or western cells containing very few or no incidents. This suggests a non-homogeneous point pattern rather than an even spatial distribution.

The kernel-smoothed intensity map confirms this, with the highest intensity in the north-east/eastern part of the city and much lower intensity elsewhere. I then aggregated incidents to wards using the coordinate-based ward assignment to identify the administrative areas contributing most to the hotspot. Maple Cross had the highest count, with 337 incidents, followed by Saffron Lea, Market End, Hazel Row and Foxley. The top five wards accounted for 70.4% of incidents and the top ten for 86.4%, confirming that incidents are highly concentrated in a small number of wards.

# Task 2

<!--- DO NOT DELETE THIS LINE - TASK 2 ANCHOR --->

Use the available neighbourhood data to develop and justify possible explanations for the patterns you identified.

```{r}
#| label: task-2-code
# Write your Task 2 code here.
```

Write your Task 2 answer here.

# Task 3

<!--- DO NOT DELETE THIS LINE - TASK 3 ANCHOR --->

Make a small number of evidence-based recommendations for priority areas or actions.

```{r}
#| label: task-3-code
# Write your Task 3 code here.
```

Write your Task 3 answer here.

# References

<!--- DO NOT DELETE THIS LINE - REFERENCES ANCHOR --->

Add references only if needed.

```{r}
#| label: word-count-check
#| echo: false
#| message: false
#| warning: false

resolve_word_count_input <- function() {
  current_input <- tryCatch(
    knitr::current_input(),
    error = function(e) NA_character_
  )

  if (!is.na(current_input) && nzchar(current_input) && file.exists(current_input)) {
    return(current_input)
  }

  if (file.exists("coursework_02.qmd")) {
    return("coursework_02.qmd")
  }

  stop(
    paste(
      "Could not locate coursework_02.qmd.",
      "Render the document from inside the coursework project folder."
    )
  )
}

extract_chunk_captions <- function(chunk_lines) {
  caption_lines <- grep(
    "^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:",
    chunk_lines,
    value = TRUE
  )

  if (length(caption_lines) == 0) {
    return(character(0))
  }

  trimws(sub("^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:\\s*", "", caption_lines))
}

count_policy_words <- function(text) {
  if (length(text) == 0) {
    return(0L)
  }

  text <- paste(text, collapse = " ")
  text <- gsub("`[^`]*`", " ", text, perl = TRUE)
  text <- gsub("\\[[^\\]]*\\]\\([^\\)]*\\)", " ", text, perl = TRUE)
  text <- gsub("https?://\\S+", " ", text, perl = TRUE)
  text <- gsub("[*_~#>]", " ", text, perl = TRUE)

  tokens <- regmatches(
    text,
    gregexpr("[[:alnum:]]+(?:[-'][[:alnum:]]+)*", text, perl = TRUE)
  )[[1]]

  if (length(tokens) == 1 && identical(tokens, character(0))) {
    return(0L)
  }

  length(tokens)
}

count_prose_words <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)

  yaml_bounds <- grep("^---\\s*$", lines)
  if (length(yaml_bounds) >= 2) {
    lines <- lines[-seq.int(yaml_bounds[1], yaml_bounds[2])]
  }

  reference_heading <- grep("^#\\s+References\\s*$", lines)
  if (length(reference_heading) > 0) {
    lines <- lines[seq_len(reference_heading[1] - 1)]
  }

  prose_lines <- character(0)
  caption_lines <- character(0)
  in_chunk <- FALSE
  current_chunk <- character(0)

  for (line in lines) {
    if (grepl("^```", line)) {
      if (!in_chunk) {
        in_chunk <- TRUE
        current_chunk <- character(0)
      } else {
        caption_lines <- c(caption_lines, extract_chunk_captions(current_chunk))
        in_chunk <- FALSE
        current_chunk <- character(0)
      }
      next
    }

    if (in_chunk) {
      current_chunk <- c(current_chunk, line)
      next
    }

    trimmed <- trimws(line)
    if (
      trimmed == "" ||
      grepl("^<!--", trimmed) ||
      grepl("^#+\\s+", trimmed)
    ) {
      next
    }

    prose_lines <- c(prose_lines, line)
  }

  total_words <- count_policy_words(prose_lines) + count_policy_words(caption_lines)
  total_words
}

true_word_count <- count_prose_words(resolve_word_count_input())
word_limit <- 1000L
status_text <- if (true_word_count <= word_limit) "under" else "over"
delta <- abs(word_limit - true_word_count)

cat(
  sprintf(
    "**Prose Word Count:** %d words (%d words %s the %d-word limit)",
    true_word_count,
    delta,
    status_text,
    word_limit
  )
)
```
